---
title: "PA1_template"
author: "mindR"
date: "April 3, 2016"
output: html_document
---

Loading and Preprocessing the data
```{r}
actdata <- read.csv("activity.csv")

#loading the data and storing it in actdata variable
```
No transformation required from the basic data analysis

Q1. What is the total number of steps taken per day?

Step 1: Calculating total number of steps taken each day

```{r}
library(dplyr)
plotdata <- actdata %>% group_by(date) %>% summarise(sum(steps, na.rm=TRUE))

colnames(plotdata) <- c("date", "TotalSteps")
```


Step 2: Creating Histogram of steps each day
```{r}
hist(plotdata$TotalSteps)
```

Step 3: calculating mean and meadian value of steps data

```{r actdata}
actmean <- mean(actdata$steps, na.rm = TRUE)
actmedian <- median(actdata$steps, na.rm = TRUE)

#mean value is 37.3826
#median value is 0
```


What is the average daily activity pattern?

Step 1: Get the daily average for each interval

```{r}
plotdata2 <- actdata %>% group_by(interval) %>% summarise(mean(steps, na.rm=TRUE))

#Stores the intervals and mean steps in the plotdata dataframe
```

Step 2: Time series plot for the steps data averages across days for each interval

```{r}
plot(plotdata2, type="l")

#plots the data and the average steps for each time interval can be seen
```

Step 3: To find the interval with maximum average steps

```{r}
#can be found using max function

#changing the column names of plotdata dataframe to make it more usable
colnames(plotdata2) <- c("interval", "meansteps")

plotdata2[plotdata2$meansteps==max(plotdata2$meansteps),]

# interval with max average steps is 835
```

Imputing missing values

Step 1: calculate and report the total missing values

```{r}
checkdata <- is.na (actdata$steps)
table(checkdata)

#total missing values 2304
```

Step 2: Strategy to fill in the missing steps values

Replacing the NAs with the equivalent mean values for each intervals

Step 3: Creating a new dataset with filled NAs for each intervals

```{r}
#function to replace the NA values and assigning it to new dataset #ModData

rep_NA <- function() {
  i <- 0
  tempData <- actdata
  for (i in which(is.na(tempData$steps))) {
    tempData$steps[i] <- plotdata2$meansteps[which(plotdata2$interval==tempData$interval[i])]
  }
  tempData
}

ModData <- rep_NA()
```

Step 4: Histogram of new dataset and checking the mean and median total number of steps and comparing it with old data

```{r}
plotdata3 <- ModData %>% group_by(date) %>% summarise(sum(steps, na.rm=TRUE))

colnames(plotdata3) <- c("date", "TotalSteps")

hist(plotdata3$TotalSteps)

#histogram seems different from the one with original data
#frequency changed for certain values because of additional data for 
#blank values

mean(ModData$steps)
median(ModData$steps)

#mean 37.3826
#median 0
#same as the original data, since missing values replaced by mean itself
```

Are there any differences in activity patterns between weekdays and weekends?

Step 1: adding a weekday column to the modified data

```{r}
#converting the date field into date class
ModData$date <- as.Date(ModData$date)

weekdays1 <- c("Monday", "Tuesday", "Wednesday", "Thursday", "Friday")

#adding a Day column 
ModData$Day <- factor((weekdays(ModData$date) %in% weekdays1), levels = c(FALSE,TRUE), labels = c("weekend", "weekday"))
```

Step 2: creating a time series plot for average number of steps averaged across weekdays and weekends

```{r}
#creating the dataframe for plotting interval vs mean steps
plotdata4 <- ModData %>% group_by(interval, Day) %>% summarise(mean(steps, na.rm=TRUE))

colnames(plotdata4) <- c("interval", "Day", "meansteps")

#creating the timesharing plot
library(lattice)
xyplot(meansteps ~ interval | Day, data = plotdata4, type = "l")
```

